---
title: "MIR Project - 750 Rock songs chord analysis"
output: html_notebook
---

Import libraries
```{r}
library(ggplot2)
library(dplyr)
library(modeest)
library(extrafont)
library(tidyr)
```



Read in the chord data for 750 songs
```{r}
complete_df <- read.csv("processed_data/750_songs_complete_chords.csv", sep = ';')
metadata_df <- read.csv("processed_data/metadata_df_total.csv", sep = ';')
metadata_df$name_title <- paste(metadata_df$artist, "-", metadata_df$title)
metadata_df$song_decade <- paste(substr(metadata_df$date, 3, 3), "0s", sep = "")
```

Join dataframes
```{r}
# Left join
joined_df <- merge(x = complete_df, y = metadata_df, by.x = "song_name", by.y = 'filename', all.x = TRUE)

joined_df$chord_duration <- joined_df$end_timestamp - joined_df$start_timestamp
joined_df$chord_simplified <- substr(joined_df$chord, 1, 2)
```

Some general description for how many songs a band has
```{r}
table(metadata_df$artist)

# metadata_df %>% 
#   count(artist) %>% 
  
bands_by_songs_and_year <- metadata_df %>% 
  count(artist) %>% 
  filter(n > 10) 

bands_by_songs_and_year$popular_decade <- c("80s", "80s", "90s", "80s", "60s", "70s", "90s", "70s", "90s", "70s", "60s", "80s", "80s", "90s", "60s", "70s", "80s", "70s")

ggplot(data = bands_by_songs_and_year, aes(y = reorder(artist, n), x = n, fill = popular_decade)) + 
  geom_bar(stat = 'identity', color = "white") +
  xlab("Count of Songs by Artist in the Top 750 Classic Rock Songs List") + 
  labs(fill = "The Artist's Most Popular Decade") +
  # coord_flip() +
  ggtitle("Led Zeppelin is the 'best' Rock Band of all-time", 
          subtitle = "According to the Top 750 Classic Rock songs list by KISW") +
  geom_text(aes(label = n), nudge_x = -1.5, size = 3, color = "white") +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_text(hjust = 0),
        plot.title = element_text(hjust = 0),
        legend.position = c(0.7, 0.4)
        )


ggsave("rock_band_songs.png", width = 8, height = 6)
```


```{r}
sapply(metadata_df, class)
```

Nr of unique chords per song
```{r}
unique_chords_df <- joined_df %>% 
  group_by(song_name) %>% 
  summarize(unique_chord_cnt = n_distinct(chord)) %>% 
  merge(y = metadata_df, by.x = "song_name", by.y = "filename", all.x = TRUE)

few_chord_song_titles <- c(
  "Baba O'riley",
  "Thunderstruck",
  # "Wish You Were Here",
  "Paranoid",
  "I Still Haven't Found What I'm Looking For",
  "Let There Be Rock",
  "Cocaine",
  "Even Flow",
  "Gimme All Your Lovin",
  "(I Can't Get No) Satisfaction",
  "Nothing Else Matters",
  "Everlong"
       )


# Cocaine
# Free Bird
# (I Can't Get No) Satisfaction
# Even Flow
# Gimme All Your Lovin
# Nothing Else Matters
# Everlong

# Stairway To Heaven
# Eruption
# Bohemian Rhapsody
# Back In Black
# We Are The Champions
# All Along The Watchtower
# Heartbreaker
# Little Wing
# Killer Queen
# Somebody To Love
# Fade To Black
# Logical Song

unique_chords_df %>% 
  filter(title %in% few_chord_song_titles) %>% 
  ggplot(aes(x = unique_chord_cnt, y = reorder(name_title, -unique_chord_cnt), fill = song_decade)) +
  geom_bar(stat = 'identity') + 
  theme_minimal() +
  scale_fill_discrete(labels = c("60s", "70s", "80s", "90s")) + 
  labs(x = " Number of Unique Chords in Song", fill = "Song Decade") + 
  geom_text(aes(label = unique_chord_cnt), nudge_x = -2, size = 4, color = "white") +
  ggtitle("A rock musician plays 3 chords for 1000 people,\nwhile a jazz musician plays 1000 chords for 3 people", 
          subtitle = "Number of Unique Chords in Popular Songs with few unique chords") + 
  xlim(c(0,10)) +
  theme(axis.title.y = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      panel.grid.major.y = element_blank(),
      panel.grid.minor.y = element_blank(),
      axis.text.x = element_blank(),
      plot.caption = element_text(hjust = 0),
      axis.title.x = element_text(hjust = 0),
      plot.title.position = "plot",
      plot.caption.position = "plot",
      # plot.title = element_text(hjust = -1),
      # legend.position = c(0.95, 0.8),
      axis.text.y = element_text(size = 9)
      )

ggsave("few_chords.png", width = 5.25, height = 6)
```


```{r}


unique_chords_df %>% 
  filter(unique_chord_cnt < 7)

# Baba O'riley
# Thunderstruck
# 309 Wish You Were Here
# Paranoid
# I Still Haven't Found What I'm Looking For
# Let There Be Rock


# 6 chords
# Cocaine
# Free Bird
# (I Can't Get No) Satisfaction
# Even Flow
# Gimme All Your Lovin
# Nothing Else Matters
# Everlong

# With many unique chords
unique_chords_df %>% 
  filter(unique_chord_cnt > 15)

# Stairway To Heaven
# Eruption
# Bohemian Rhapsody
# Back In Black
# We Are The Champions
# All Along The Watchtower
# Heartbreaker
# Little Wing
# Killer Queen
# Somebody To Love
# Fade To Black
# Logical Song



many_chord_song_titles <- c(
  "Stairway To Heaven",
  "Eruption",
  "Bohemian Rhapsody",
  "Back In Black",
  "We Are The Champions",
  "All Along The Watchtower",
  "Heartbreaker",
  "Little Wing",
  "Killer Queen",
  "Somebody To Love",
  "Fade To Black",
  "Logical Song"
)


unique_chords_df %>% 
  filter(title %in% many_chord_song_titles) %>% 
  ggplot(aes(x = unique_chord_cnt, y = reorder(name_title, unique_chord_cnt), fill = song_decade)) +
  geom_bar(stat = 'identity') + 
  theme_minimal() +
  scale_fill_discrete(labels = c("60s", "70s", "80s", "90s")) + 
  labs(x = " Number of Unique Chords in Song", fill = "Song Decade") + 
  geom_text(aes(label = unique_chord_cnt), nudge_x = -3, size = 4, color = "white") +
  ggtitle("Complicated songs also get a Whole Lotta Love", 
          subtitle = "Number of Unique Chords in Popular Songs with few unique chords") + 
  theme(axis.title.y = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      panel.grid.major.y = element_blank(),
      panel.grid.minor.y = element_blank(),
      axis.text.x = element_blank(),
      # plot.title = element_text(hjust = -1),
      plot.caption = element_text(hjust = -1),
      plot.title.position = "plot",
      plot.caption.position = "plot",
      axis.title.x = element_text(hjust = 0),
      # legend.position = c(0.95, 0.8),
      axis.text.y = element_text(size = 10) 
      )


ggsave("many_chords.png", width = 5.25, height = 6)
```

Linegraph for mean number of unique chords per year
```{r}
# https://www.rockmusictimeline.com/1985

events_labels <- data.frame(
  date = c(1982, 1985, 1992, 1999, 1970, 1972),
  avg_unique_chord_cnt_per_year = c(14.3, 6, 13, 6.5, 14.5, 7),
  label = c("1981 - MTV is established", 
            "1985 - Live Aid concert", 
            "1992 - Nirvana's\n'Smells Like Teen Spirit'", 
            "1999 - Napster is launched",
            "1967 - Beatle's Sgt Pepper's\n and Jimi Hendrix Experience",
            "1972 - Rise of Glam Rock")
)

unique_chords_df %>% 
  group_by(date) %>% 
  summarize(avg_unique_chord_cnt_per_year = mean(unique_chord_cnt), decade = paste(substr(date, 3,3), "0s", sep = "")) %>% 
  ggplot(aes(x = date, y = avg_unique_chord_cnt_per_year)) +
  geom_line(aes(color = decade), size = 0.5) +
  geom_point(aes(color = decade) ,size = 1.5) +
  # geom_area(alpha = 0.3) +
  theme_minimal() +
  ylim(c(0,18)) +
  xlim(c(1967,2005)) +
  ylab("Average Number of Chords in Songs") +
  xlab("Year") + 
  labs(color = "Decade") +
  # geom_text(data = events_labels, aes(label = label), nudge_y = 1.5, size = 3) +
  geom_label(data = events_labels, aes(label = label), nudge_y = 0.5, size = 3) +
  scale_color_discrete(breaks = c("60s", "70s", "80s", "90s", "00s")) + 
  # scale_color_brewer(palette="Greens") +
  ggtitle("Creativity of Rock Bands Might've Peaked in the 80s", 
        subtitle = "Average Number of Unique Chords for Songs Made in a Particular Year") + 
  theme(axis.title.y = element_blank(),
    # panel.grid.major.x = element_blank(),
    # panel.grid.minor.x = element_blank(),
    # panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    # axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    plot.title = element_text(hjust = 0),
    legend.position = c(0.9, 0.86)
    )

ggsave("unique_chords_year_plot.png", width = 8, height = 6)
```


Barplot with most common chords (by total time)
With radial coordinates to signify the circle of fifths!
```{r}

sum(joined_df$chord_duration) # 63336.67

joined_df %>% 
  group_by(chord_simplified) %>% 
  summarize(total_chord_length = sum(chord_duration),
            total_chord_proportion = sum(chord_duration)/63336.67) %>% 
  filter(chord_simplified != "N") %>% 
  ggplot(aes(x = factor(chord_simplified, 
                        level = c("G:", "D:", "A:", "E:", 
                                  "B:", "Gb", "Db", "Ab", 
                                  "Eb", "Bb", "F:", "C:")), 
             y = total_chord_proportion)) +
  geom_bar(aes(fill = total_chord_proportion), stat = 'identity') +
  # scale_fill_continuous(type = "viridis") + 
  # scale_fill_viridis_c(option = "magma") + 
  scale_fill_gradient(low = "#003f5c",
                      high = "#ff6361", 
                      name = "Chord Proportion in\nRock Song Playtime",
                      labels = scales::percent) + 
  # scale_colour_brewer(palette = "Set1") +
  scale_x_discrete(labels = c("G", "D", "A", "E", 
                              "B", "Gb", "Db", "Ab", 
                              "Eb", "Bb", "F", "C")) +
  coord_polar() + 
  ggtitle('Guitarists Prefer Easier Chords',
          subtitle = "Time spent using chords in the Circle of Fifths") + 
  theme_minimal() +
  theme(axis.title.y = element_blank(),
  # panel.grid.major.x = element_blank(),
  # panel.grid.minor.x = element_blank(),
  # panel.grid.major.y = element_blank(),
  panel.grid.minor.y = element_blank(),
  axis.text.y = element_blank(),
  axis.title.x = element_blank(),
  plot.title = element_text(hjust = 0),
  axis.text.x = element_text(face = "bold", size = 12, vjust = 15) 
  )
  

# joined_df %>% 
#   group_by(chord) %>% 
#   summarize(total_chord_length = sum(chord_duration)) %>% 
#   filter(total_chord_length > 1000) %>% 
#   ggplot(aes(x = chord, y = total_chord_length)) +
#   geom_bar(stat = 'identity') 

ggsave("radial_plot.png", width = 6, height = 6)
```
```{r}

artists_vector <- c("Led Zeppelin", "Jimi Hendrix", "AC/DC", "Van Halen", 
              "Aerosmith", "Pink Floyd", "Metallica", "Pearl Jam",
             "Def Leppard", "Ozzy Osbourne", "The Who", "Rolling Stones",
             "Soundgarden", "The Doors", "Alice In Chains"
             )

bands_chords_type_df <- joined_df %>%
  filter(artist %in% artists_vector) %>% 
  group_by(chord_simplified, artist) %>% 
  summarize(total_chord_length = sum(chord_duration),
            total_chord_proportion = sum(chord_duration)/63336.67) %>% 
  filter(chord_simplified != "N") 

band_major_chords <- joined_df %>% 
  filter(artist %in% artists_vector) %>% 
  filter(grepl("maj", chord)) %>% 
  group_by(chord_simplified, artist) %>% 
  summarize(total_major_length = sum(chord_duration))

band_minor_chords <- joined_df %>% 
  filter(artist %in% artists_vector) %>% 
  filter(grepl("min", chord)) %>% 
  group_by(chord_simplified, artist) %>% 
  summarize(total_minor_length = sum(chord_duration))


bands_chords_type_df <- merge(bands_chords_type_df, band_major_chords, by = c("chord_simplified", "artist"), all.x = TRUE)
bands_chords_type_df <- merge(bands_chords_type_df, band_minor_chords, by = c("chord_simplified", "artist"), all.x = TRUE)

bands_chords_type_df$total_chord_length_for_band <- ave(bands_chords_type_df$total_chord_length, bands_chords_type_df$artist, FUN = sum)

bands_chords_type_df$total_chord_proportion <- bands_chords_type_df$total_chord_length/bands_chords_type_df$total_chord_length_for_band

bands_chords_type_df$total_neutral_length <- bands_chords_type_df$total_chord_length - bands_chords_type_df$total_major_length - bands_chords_type_df$total_minor_length

bands_chords_type_df$total_major_proportion <- bands_chords_type_df$total_major_length/bands_chords_type_df$total_chord_length_for_band

bands_chords_type_df$total_minor_proportion <- bands_chords_type_df$total_minor_length/bands_chords_type_df$total_chord_length_for_band

bands_chords_type_df$total_neutral_proportion <- bands_chords_type_df$total_neutral_length/bands_chords_type_df$total_chord_length_for_band


bands_chords_type_df_tall <- gather(bands_chords_type_df, key, value, total_major_proportion, total_minor_proportion, total_neutral_proportion)

# all_chords_tall <- gather(all_chords, key, value, total_neutral_length, total_major_length, total_minor_length)
  
# bands_chords_type_df_tall %>% 
#   ggplot(aes(x = factor(chord_simplified, 
#                         level = c("G:", "D:", "A:", "E:", 
#                                   "B:", "Gb", "Db", "Ab", 
#                                   "Eb", "Bb", "F:", "C:")), 
#              y = total_chord_proportion)) +
#   geom_bar(aes(fill = total_chord_proportion), stat = 'identity') +
#   # scale_fill_continuous(type = "viridis") + 
#   # scale_fill_viridis_c(option = "magma") + 
#   scale_fill_gradient(low = "#003f5c",
#                       high = "#ff6361", 
#                       name = "Chord Proportion in\nRock Song Playtime",
#                       labels = scales::percent) + 
#   # scale_colour_brewer(palette = "Set1") +
#   scale_x_discrete(labels = c("G", "D", "A", "E", 
#                               "B", "Gb", "Db", "Ab", 
#                               "Eb", "Bb", "F", "C")) +
#   coord_polar() + 
#   ggtitle('Guitarists Prefer Easier Chords',
#           subtitle = "Time spent using chords in the Circle of Fifths") + 
#   theme_minimal() +
#   theme(axis.title.y = element_blank(),
#   # panel.grid.major.x = element_blank(),
#   # panel.grid.minor.x = element_blank(),
#   # panel.grid.major.y = element_blank(),
#   panel.grid.minor.y = element_blank(),
#   axis.text.y = element_blank(),
#   axis.title.x = element_blank(),
#   plot.title = element_text(hjust = 0),
#   axis.text.x = element_text(face = "bold", size = 12, vjust = 15) 
#   ) +
#   facet_wrap(~artist, nrow = 2, ncol = 4)
#   # facet_grid(~artist, rows = 2, cols = 4, space = "fixed", scales = "fixed")

bands_chords_type_df_tall %>% 
  ggplot(aes(x = factor(chord_simplified, 
                        level = c("G:", "D:", "A:", "E:", 
                                  "B:", "Gb", "Db", "Ab", 
                                  "Eb", "Bb", "F:", "C:")), 
             y = value, 
             fill = key)) +
  geom_bar(stat = 'identity', position = position_stack(reverse = TRUE))  +
  scale_fill_manual(name = "Chord Mode", 
                      labels = c("Major", "Minor", "Neutral"),
                      values = c("#ffa600", "#ff6361", "#003f5c")) +
                      # values = c("#003f5c", "#ff6361", "#ffa600")) +
                      # values = c("#ff6361", "#ffa600", "#003f5c")) +
  scale_x_discrete(labels = c("G", "D", "A", "E", 
                              "B", "Gb", "Db", "Ab", 
                              "Eb", "Bb", "F", "C")) +
  coord_polar() + 
  ggtitle('Rock Bands Prefer Chords That Are Easier To Play On Guitar',
          subtitle = "The playtime of a chord and its mode in proportion to the whole band's playtime") + 
  theme_minimal() +
  theme(axis.title.y = element_blank(),
  # panel.grid.major.x = element_blank(),
  # panel.grid.minor.x = element_blank(),
  # panel.grid.major.y = element_blank(),
  panel.grid.minor.y = element_blank(),
  axis.text.y = element_blank(),
  axis.title.x = element_blank(),
  plot.title = element_text(hjust = 0, size = 24),
  plot.subtitle = element_text(hjust = 0, size = 16, margin=margin(0,0,20,0)),
  axis.text.x = element_text(face = "bold", size = 10, vjust = 15),
  strip.text = element_text(size = 16), # facet title size
  panel.spacing = unit(1.5, "lines"), # facet spacing
  legend.key.size = unit(2, "lines"),
  legend.title = element_text(size = 24),
  legend.text = element_text(size = 20),
  legend.position = c(1.2, 0.9)
  ) +
  facet_wrap(~factor(artist, levels = c("Soundgarden", "The Who", "Pink Floyd", "Pearl Jam", "Led Zeppelin",
                                         "Jimi Hendrix", "Metallica", "Aerosmith", "Def Leppard", "AC/DC",
                                        "Van Halen", "Alice In Chains", "Ozzy Osbourne", "Rolling Stones", "The Doors")), 
             nrow = 3, 
             ncol = 5)

ggsave("bands_circle_of_fifths.png", width = 14, height = 8)
```

```{r}
joined_df %>% 
  group_by(chord_simplified) %>% 
  summarize(total_chord_length = sum(chord_duration),
            total_chord_proportion = sum(chord_duration)/63336.67) %>% 
  filter(chord_simplified != "N")
```


Minor and Major and Neutral
Lineplot vs Year
or barplot with fill to 1
```{r}
major_chords <- joined_df %>% 
  filter(grepl("maj", chord)) %>% 
  group_by(date) %>% 
  summarize(total_major_length = sum(chord_duration))

minor_chords <- joined_df %>% 
  filter(grepl("min", chord)) %>% 
  group_by(date) %>% 
  summarize(total_minor_length = sum(chord_duration))

all_chords <- joined_df %>% 
  # filter(grepl("maj", chord)) %>% 
  group_by(date) %>% 
  summarize(total_chord_length = sum(chord_duration))

all_chords <- merge(all_chords, major_chords, by = "date", all.x = TRUE)
all_chords <- merge(all_chords, minor_chords, by = "date", all.x = TRUE)

all_chords$total_neutral_length <- all_chords$total_chord_length - (all_chords$total_major_length + all_chords$total_minor_length)

# all_chords$major_proportion <- all_chords$total_major_length/all_chords$total_chord_length
# all_chords$minor_proportion <- all_chords$total_minor_length/all_chords$total_chord_length
# all_chords$neutral_proportion <- all_chords$total_neutral_length/all_chords$total_chord_length

all_chords_tall <- gather(all_chords, key, value, total_neutral_length, total_major_length, total_minor_length)

all_chords_tall %>% 
  ggplot(aes(x = date, y = value, fill = key)) +
  geom_bar(position = "fill", stat = "identity") +
  xlim(c(1968, 2004)) +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(breaks = seq(1968, 2000, 2),
                   limits = c(1968, 2002)) +
  scale_fill_manual(name = "Chord Mode", 
                      labels = c("Major", "Minor", "Neutral"),
                      # values = c("#ffa600", "#ff6361", "#003f5c")) +
                      values = c("#003f5c", "#ff6361", "#ffa600")) +
  ggtitle('Rock is Surprisingly Happy',
        subtitle = "Proportions of Time Spent Using Different Chord Modes") + 
  theme_minimal() +
  theme(axis.title.y = element_blank(),
  panel.grid.major.x = element_blank(),
  panel.grid.minor.x = element_blank(),
  panel.grid.major.y = element_blank(),
  panel.grid.minor.y = element_blank(),
  axis.text.y = element_blank(),
  axis.title.x = element_blank(),
  axis.text.x = element_text(angle = 90, hjust = 0, vjust = 2, size = 10),
  )

ggsave("chord_mode_year.png", width = 8, height = 6)
```



Heatmap for chords and years

```{r}
joined_df %>% 
  group_by(chord_simplified, date) %>% 
  summarize(total_chord_length = sum(chord_duration),
            total_chord_proportion = sum(total_chord_length)/sum(total_chord_length)) %>% 
  filter(chord_simplified != "N") %>% 
  ggplot(aes(x = date, y = chord_simplified, fill = total_chord_proportion)) +
  geom_tile() +
  xlim(c(1968, 2005)) +
  theme_minimal()


chord_length_year <- joined_df %>% 
  group_by(chord_simplified, date) %>% 
  summarize(total_chord_length = sum(chord_duration))

year_total_chord_length <- joined_df %>% 
  group_by(date) %>% 
  summarize(total_length = sum(chord_duration))

merged_df <- merge(chord_length_year, year_total_chord_length, by = "date")
merged_df$total_chord_proportion <- merged_df$total_chord_length/merged_df$total_length

merged_df %>% 
  filter(chord_simplified != "N") %>% 
  ggplot(aes(x = date, 
             y = factor(chord_simplified,
                        levels = c("Ab", "A:", "Bb", "B:",
                                   "C:", "Db", "D:", "Eb",
                                   "E:", "F:", "Gb", "G:")), 
             fill = total_chord_proportion)) +
  geom_tile(alpha = 1) +
  # scale_fill_distiller(type = "Greens") +
  # scale_fill_continuous(type = "viridis") + 
  scale_fill_gradient2(low = "#003f5c", 
                        mid = "#bc5090",
                        high = "#ffa600", 
                       space = "Lab",
                       midpoint = 0.35,
                       name = "Chord Proportion of Total\nRock Songs Playtime ",
                       labels = scales::percent) + 
  scale_y_discrete(labels = c(
                    "Ab", "A", "Bb", "B",
                    "C", "Db", "D", "Eb",
                    "E", "F", "Gb", "G"
    
            )) +
  # xlim(c(1968, 2005)) +
  scale_x_continuous(breaks = seq(1968, 2000, 2),
                     limits = c(1968, 2002)) +
  theme_minimal() +
  ggtitle("E and A are all-time favorites",
          subtitle = "Prevalence of chords during the Age of Rock") +
  theme(axis.title.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    axis.title.x = element_blank(),
    plot.title = element_text(hjust = 0),
    axis.text.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(angle = 90, hjust = 0, vjust = 1.5, size = 10),
    panel.background = element_rect(fill = "#003f5c")
  )

ggsave("chord_year_heatmap.png", width = 10, height = 6)
```


